Adaptive Algorithms Employing Continuous Linear Functionals

### 2018-08-04 AutoAlgorithmLinearFunctionalDataVerC.tex
revised proof in Section 4

The proof is accomplished by constructing fooling functions in $\cc$ with norms no greater than $\varrho$.

\begin{theorem} \label{thm:lowbdcomp}
A lower bound on the complexity of the linear problem defined in \eqref{probDef} is
\begin{subequations} \label{compbd}
\begin{align}
 \label{compbdA}
\comp(&\ca(\cc),\varepsilon,\rho) \ge n_{j^*},  \\
\nonumber
\text{where} & \\
\label{compbdB}
& j^* = \max \left \{ j \in \naturals : \left(1+\frac{(a-1)^2}{(a+1)^{2} A^2 }\right)\sum_{k=1}^j \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}   < 
\frac{(a-1)^2\rho^2}{(a+1)^2A^2\varepsilon^2}
\right \}.
\end{align}
\end{subequations}
\end{theorem}

\begin{proof}
Consider a fixed $\rho$ and $\varepsilon$.  Choose any positive integer $j$ such that $n_j$ exceeds $\comp(\ca(\cc),\varepsilon,\rho)$.  The proof proceeds by constructing test input functions, $f$ and $f^*_{\pm}$, all lying in $\cc$ and in the ball with radius $\rho$, which yield the same approximate solution while yielding different true solutions.  This leads to a lower bound on $n_j$, which can be translated into a lower bound on $\comp(\ca(\cc),\varepsilon,\rho)$.

The first test function $f \in \cc$ is defined in terms of its series coefficients:
\begin{align}
\nonumber
\hf_i &:= \begin{cases}
\displaystyle
\frac{c b^{k-j}}{\lambda_{n_{k}}},  & i =  n_{k}, \ k = 1, \ldots, j,
\\
0, & \text{otherwise}
\end{cases}
\\
\label{cdef}
c^2 &:=  \rho^2 \left[\left(1+\frac{(a-1)^2}{(a+1)^{2} A^2 }\right)\sum_{k=1}^j \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}\right]^{-1},
\end{align}
where 
\[A = \max_{1 \le k \le j} \frac{\lambda_{n_{k-1}+1}}{\lambda_{n_k}}.\]
It can be verified that the test function lies in a ball of radius $\rho$ and also in the cone $\cc$:
\begin{align}
\label{testfunnorm}
\norm[\cf]{f}^2 &=  c^2\sum_{k=1}^j \frac{ b^{2(k-j)}}{\lambda^2_{n_{k}}} \le \rho^2,
\\
\nonumber
\sigma_k(f) &= \begin{cases}
\displaystyle
c b^{k-j}, & k =1, \ldots, j, \\
0, & \text{otherwise},
\end{cases}
\\
\nonumber
\sigma_{k+r}(f) &= \begin{cases}
\displaystyle 
b^{r} \sigma_k(f) \le a b^r \sigma_k(f), & k+r \le j, \ r \ge 1,
\\
0 \le a b^r \sigma_k(f), & k+r > j, \ r \ge 1.
\end{cases}
\end{align}

Now suppose that $A^* \in \ca(\cc)$ is an optimal algorithm, i.e., $\cost(A^*,\cc,\varepsilon,\rho) =  \comp(\ca(\cc),\varepsilon,\rho)$ for all $\varepsilon, \rho > 0$.  For our particular input $f$ defined above, suppose that $A^*(f,\varepsilon)$ samples $L_1(f), \ldots, L_n(f)$ where 
\[n + 1
\le \comp(\ca(\cc),\varepsilon,\rho) +1 < n_j.\]  Let $u$ be a linear combination of $u_1, \cdots, u_{n_j}$ constructed to have unit norm and satisfy $L_1(u) = \cdots = L_n(u) = 0$ and $\langle u,f \rangle_{\cf} = 0$. Denote
\[u = \sum_{k=1}^{j}u^{(k)}\frac{b^{k-j}}{\lambda_{n_k}}\]
where $u^{k}= \sum_{i = n_{k-1}+1}^{n_k} \hu_i u_i$ and
$\max_{1\le k\le j} \norm[\cf]{u^{(k)}} = 1.$
Let 
\begin{gather}
\label{etadef}
f^*_\pm = f \pm \eta u, \quad \eta : =  \frac {(a-1)}{(a+1)A}c, \\
\nonumber
\norm[\cf]{f^*_\pm}^2 \le \norm[\cf]{f}^2 + \norm[\cf]{\eta u } 
\le \sum_{k=1}^j \frac{ b^{2(k-j)}}{\lambda^2_{n_{k}}} \left(c^2+ \eta^2 \norm[\cf]{u^{(k)}}^2\right)\\
\nonumber
\le \sum_{k=1}^j \frac{ b^{2(k-j)}}{\lambda^2_{n_{k}}} \left(c^2+ \eta^2 \right)  \le \rho^2.
\end{gather} 
By design, $A^*(f^*_\pm,\varepsilon) = A^*(f,\varepsilon)$, which will be used below.

Now we must check that $f^*_\pm \in \cc$. From the definition in \eqref{sumdef} it follows that for $k = 1, \ldots, j$ and $r \ge 1$
\begin{align*}
\sigma_k(f^*_\pm) & \begin{cases} 
\displaystyle
\le \sigma_k(f) + \sigma_k(\eta u)\le 
c b^{k-j} + \eta\lambda_{n_{k-1}+1}\frac{b^{k-j}\norm[\cf]{u^{(k)} } }{\lambda_{n_k}}
\le b^{k-j}\left(c+\eta A \right) 
\\[1ex]
\displaystyle
\ge \sigma_k(f) - \sigma_k(\eta u)\ge 
c b^{k-j} - \eta\lambda_{n_{k-1}+1}\frac{b^{k-j}\norm[\cf]{u^{(k)} } }{\lambda_{n_k}}
\ge b^{k-j}\left(c-\eta A \right) , 
\end{cases}
\\
\sigma_{k+r}(f^*_\pm)&
\le b^{k+r-j}(c+\eta A) = ab^r b^{k-j}\frac{2c}{a+1}
=ab^r b^{k-j}\left(c-\eta A \right) \le a b^r \sigma_{k}(f^*_\pm).
\end{align*}

Given the two test functions $f^*_\pm$ and our successful algorithm, $A^*$, we note that
\begin{align*}
\varepsilon &\ge \max \bigl\{\norm[\cg]{S(f^*_+) - A^*(f^*_+,\varepsilon)}, \norm[\cg]{S(f^*_-) - A^*(f^*_-,\varepsilon)} \bigr\} \\
&\ge \frac 12 \bigl[\norm[\cg]{S(f^*_+) - A^*(f,\varepsilon)} + \norm[\cg]{S(f^*_-) - A^*(f,\varepsilon)}  \bigr] \\
&\qquad \qquad \text{since } A^*(f^*_\pm,\varepsilon) = A^*(f,\varepsilon) \\
&\ge \frac 12 \norm[\cg]{S(f^*_+) - S(f^*_-)} \quad \text{by the triangle inequality}\\
&\ge \frac 12 \norm[\cg]{S(f^*_+ - f^*_-)} \quad \text{since $S$ is linear}\\
&= \eta \norm[\cg]{S(u)}
\end{align*}
Thus, we have
\begin{align*}
\varepsilon^2 \ge \eta \norm[\cg]{S(u)}^2 & = 
\eta^2\sum_{k=1}^{j} \norm[\cf]{S(u^{(k)}}^2 
\left(\frac{b^{k-j}}{\lambda_{n_k}}\right)^2\\
& \ge \eta^2 
\sum_{k=1}^{j} {\sigma_k(u)}^2 
\left(\frac{b^{k-j}}{\lambda_{n_k}}\right)^2\\
& \ge \eta^2 
\sum_{k=1}^{j} \norm[\cf]{u^{(k)}}^2 
b^{2(k-j)}\\
& \ge  \eta^2 b^{2(k^{*}-j)} \qquad \text{ where } k^* = \arg\max_{k} \norm[\cf]{u^{(k)}} = 1\\
& \ge \eta^2 = \frac{(a-1)^2}{(a+1)^2A^2}c^2 \\
&
=\frac{(a-1)^2 \rho^2}{(a+1)^2A^2}\left[\left(1+\frac{(a-1)^2}{(a+1)^{2} A^2 }\right)\sum_{k=1}^j \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}\right]^{-1}
\end{align*}

This lower bound must be satisfied by $j$ to be consistent with the assumption $\comp(\ca(\cc),\varepsilon,\rho) < n_j$.  Any $j$ violating this inequality then provides a lower bound on $\comp(\ca(\cc),\varepsilon,\rho)$.
\end{proof}

\begin{theorem}
If the $\lambda_{n_j}$ satisfy the lower bound
\begin{equation} \label{lambdaalphabetabd}
\lambda_{n_j} \ge \xi \zeta^{j}, \qquad j \in \naturals,
\end{equation}
for some non-negative $\xi$ and $\zeta$ with $\zeta < 1$, then $\cost(\tA,\cc,\varepsilon,\rho)$ is essentially the same as $\comp(\ca(\cc),\varepsilon,\rho)$ for all positive $\varepsilon$ and $\rho$.  Thus, our adaptive algorithm, $\tA$ is essentially optimal for problems defined on $\cc$.
\end{theorem}
\begin{proof}  Fix positive $\varepsilon$ and $\rho$.  Without loss of generality, we may choose $\zeta$ in \eqref{lambdaalphabetabd} different from $b$.  Let $j^*$ be defined as in \eqref{compbdB}, and consider the sum in its definition.  For all $j \in \naturals$, 
\begin{align*}
\sum_{k=1}^j \frac{b^{k-j}}{\lambda_{n_{k}}}
& \le 
\frac{1}{\xi b^j} \sum_{k=1}^j \left(\frac{b}{\zeta} \right)^{k-j} 
= 
\frac{1}{\xi b^j} \frac{b/\zeta - (b/\zeta)^{j+1}}{1 - b/\zeta}
\\
&\le
\begin{cases}
\displaystyle 
\frac{b}{\xi \zeta^{j}(b -  \zeta)},
& \zeta < b
\\
\displaystyle 
\frac{b}{\xi b^{j}(\zeta - b)}, & \zeta > b.
\end{cases}
\end{align*}

Let $n_{j^\dagger} = \cost(\tA,\cc,\varepsilon,\rho)$.  We consider two cases.

If $\zeta > b$, then 
\begin{align*}
\sum_{k=1}^{j^\dagger} \frac{b^{k-j^\dagger}}{\lambda_{n_{k}}}
& \le \frac{b}{\xi b^{j^\dagger}(\zeta - b)} \\ 
& < \frac{\rho a^2 b \lambda_{n_0+1}}{\varepsilon \xi \sqrt{1-b^2}(\zeta - b) } \\
& = \frac{\rho (a-1)}{2 a (\omega\varepsilon)}, \quad \text{where } \omega = \frac{\xi (a-1)\sqrt{1-b^2}(\zeta - b)}{2a^3b \lambda_{n_0+1}}  
\end{align*}
Thus, by \eqref{compbd}, $\cost(\tA,\cc,\varepsilon,\rho) = n_{j^\dagger} \le \comp(\ca(\cc),\omega \varepsilon,\rho)$.

Alternatively, if $\zeta < b$, then 
\begin{align*}
\sum_{k=1}^{j^\dagger} \frac{b^{k-j^\dagger}}{\lambda_{n_{k}}}
& \le \frac{b}{\xi \zeta^{j^\dagger}(b - \zeta)} \\ 
& < \frac{\rho a^2 b \lambda_{n_0+1}}{\varepsilon \xi \sqrt{1-b^2}(\zeta - b) } \\
& = \frac{\rho (a-1)}{2 a (\omega\varepsilon)}, \quad \text{where } \omega = \frac{\xi (a-1)\sqrt{1-b^2}(\zeta - b)}{2a^3b \lambda_{n_0+1}}  
\end{align*}

\end{proof}

@@

\yuhannote{The figure is here in comment}
%\begin{figure}
%	\centering
%	\includegraphics[width = 8 cm]{ProgramsImages/SimDirectSolErr.eps}
%\end{figure}

>>>>

A lower bound on $\comp(\ca(\cc),\varepsilon,\rho)$ is established by   proof is accomplished by constructing fooling functions in $\cc$ with norms no greater than $\rho$.  To obtain a result that can be compared with the cost of our algorithm, we assume that 
\begin{equation} \label{lambdaRatio}
R = \sup_{k \in \naturals} \frac{\lambda_{n_{k-1}}}{\lambda_{n_k}} < \infty.
\end{equation}
This means that the $n_k$ are not too far apart with respect to the decay of $\lambda_i$ as $i \to \infty$

\begin{theorem} \label{thm:lowbdcomp}
Under assumption  \eqref{lambdaRatio}, a lower bound on the complexity of the linear problem defined in \eqref{probDef} is
\begin{subequations} \label{compbd}
\begin{align}
 \label{compbdA}
&\comp(\ca(\cc),\varepsilon,\rho) \ge n_{j^*}, 
\intertext{where}
\label{compbdB}
j^* & = \max \left \{ j \in \naturals : \left[\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right] \sum_{k=0}^j \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}   <
\frac{\rho^2}{\varepsilon^2}
\right \}.
\end{align}
\end{subequations}
\end{theorem}

\begin{proof}
Consider a fixed $\rho$ and $\varepsilon$.  Choose any positive integer $j$ such that $n_j$ exceeds $\comp(\ca(\cc),\varepsilon,\rho)$.  The proof proceeds by carefully constructing thre  test input functions, $f$ and $f_{\pm}$, lying in $\cc \cap \cb_{\rho}$, which yield the same approximate solution but different true solutions.  This leads to a lower bound on $n_j$, which can be translated into a lower bound on $\comp(\ca(\cc),\varepsilon,\rho)$.

The first test function $f \in \cc$ is defined in terms of its series coefficients:
\begin{align}
\nonumber
\hf_i &:= \begin{cases}
\displaystyle
\frac{c b^{k-j}}{\lambda_{n_{k}}},  & i =  n_{k}, \ k = 1, \ldots, j,
\\
0, & \text{otherwise},
\end{cases}
\\
\label{cdef}
c^2 &:=  \rho^2 \left[ \left(1+\frac{(a-1)^2}{(a+1)^{2} R^2 }\right)\sum_{k=0}^j \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}\right]^{-1}.
\end{align}
It can be verified that the test function lies both in $\cb_{\rho}$ and in $\cc$:
\begin{align}
\label{testfunnorm}
\norm[\cf]{f}^2 &=  c^2\sum_{k=1}^j \frac{ b^{2(k-j)}}{\lambda^2_{n_{k}}} \le \rho^2,
\\
\nonumber
\sigma_k(f) &= \begin{cases}
\displaystyle
c b^{k-j}, & k =1, \ldots, j, \\
0, & \text{otherwise},
\end{cases}
\\
\nonumber
\sigma_{k+r}(f) &= \begin{cases}
\displaystyle 
b^{r} \sigma_k(f) \le a b^r \sigma_k(f), & k+r \le j, \ r \ge 1,
\\
0 \le a b^r \sigma_k(f), & k+r > j, \ r \ge 1.
\end{cases}
\end{align}

Now suppose that $A^* \in \ca(\cc)$ is an optimal algorithm, i.e., $\cost(A^*,\cc,\varepsilon,\rho) =  \comp(\ca(\cc),\varepsilon,\rho)$ for all $\varepsilon, \rho > 0$.  For our particular input $f$ defined above, suppose that $A^*(f,\varepsilon)$ samples $L_1(f), \ldots, L_n(f)$ where 
\[
n + 1\le \comp(\ca(\cc),\varepsilon,\rho) +1 < n_j.
\]

Let $u$ be a linear combination of $u_1, \cdots, u_{n_j}$, expressed as
\[
u =  \sum_{k=0}^{j}\frac{b^{k-j}u^{(k)}}{\lambda_{n_k}},
\]
where $u^{(0)}$ is a linear combination of $u_{1}, \ldots, u_{n_0}$, and each $u^{(k)}$ is a linear combination of $u_{n_k+1}, \ldots, u_{n_k}$, for $k =1, \ldots, j$.  We constrain $u$ to satisfy:
\begin{equation}\label{uConstraint}
L_1(u) = \cdots = L_n(u) = 0, \qquad \langle u,f \rangle_{\cf} = 0, \qquad 
\max_{0\le k\le j} \bignorm[\cf]{u^{(k)}} = 1.
\end{equation}
Since $u$ is a linear combination of $n_j >n+1$ basis functions, these constraints can be satisfied.

Let the other two test functions be constructed in terms of $u$ as 
\begin{align}
\label{etadef}
f_\pm & := f \pm \eta u, \qquad \eta : =  \frac {(a-1)c}{(a+1)R}, \\
\nonumber
\norm[\cf]{f^*_\pm}^2 & \le \norm[\cf]{f}^2 + \norm[\cf]{\eta u }^2 \qquad \text{by \eqref{uConstraint}} \\
& \le \sum_{k=1}^j \frac{ b^{2(k-j)}}{\lambda^2_{n_{k}}} \left(c^2+ \eta^2 \bignorm[\cf]{u^{(k)}}^2\right) + \eta^2 \bignorm[\cf]{u^{(0)}}^2 \frac{b^{-2j}}{\lambda_{n_0}^2}\\
\nonumber
& \le  \left(c^2+ \eta^2 \right) \sum_{k=0}^j \frac{ b^{2(k-j)}}{\lambda^2_{n_{k}}}  \qquad \text{by \eqref{uConstraint}}\\
&  \le \rho^2,
\end{align} 
so $f_{\pm} \in \cb_{\rho}$.  By design, $A^*(f_\pm,\varepsilon) = A^*(f,\varepsilon)$, which will be used below.

Now we must check that $f_\pm \in \cc$. From the definition in \eqref{sumdef} it follows that for $k = 1, \ldots, j$ and $r \ge 1$
\begin{equation*}
\sigma_k(f_\pm)  \begin{cases} 
\displaystyle
\le \sigma_k(f) + \sigma_k(\eta u)\le 
c b^{k-j} + \eta\lambda_{n_{k-1}+1}\frac{b^{k-j}\norm[\cf]{u^{(k)} } }{\lambda_{n_k}}
\le b^{k-j}\left(c+\eta R \right) 
\\[1ex]
\displaystyle
\ge \sigma_k(f) - \sigma_k(\eta u)\ge 
c b^{k-j} - \eta\lambda_{n_{k-1}+1}\frac{b^{k-j}\norm[\cf]{u^{(k)} } }{\lambda_{n_k}}
\ge b^{k-j}\left(c-\eta R \right) , 
\end{cases}
\end{equation*}
Therefore, 
\begin{equation*}
\sigma_{k+r}(f_\pm)
\le b^{k+r-j}(c+\eta R) = ab^r b^{k-j}\frac{2c}{a+1}
=ab^r b^{k-j}\left(c-\eta R \right) \le a b^r \sigma_{k}(f_\pm),
\end{equation*}
which establishes that $f_\pm \in \cc$.

Although two test functions $f_\pm$ yield the same approximate solution, they have different true solutions.  In particular,
\begin{align*}
\varepsilon &\ge \max \bigl\{\norm[\cg]{S(f_+) - A^*(f_+,\varepsilon)}, \norm[\cg]{S(f_-) - A^*(f_-,\varepsilon)} \bigr\} \\
&\ge \frac 12 \bigl[\norm[\cg]{S(f^*_+) - A^*(f,\varepsilon)} + \norm[\cg]{S(f^*_-) - A^*(f,\varepsilon)}  \bigr] \\
&\qquad \qquad \text{since } A^*(f^*_\pm,\varepsilon) = A^*(f,\varepsilon) \\
&\ge \frac 12 \norm[\cg]{S(f^*_+) - S(f^*_-)} \quad \text{by the triangle inequality}\\
&\ge \frac 12 \norm[\cg]{S(f^*_+ - f^*_-)} \quad \text{since $S$ is linear}\\
&= \eta \norm[\cg]{S(u)}.
\end{align*}
Thus, we have
\begin{align*}
\varepsilon^2  & \ge \eta \norm[\cg]{S(u)}^2= 
\eta^2 \sum_{k=0}^{j} \bignorm[\cg]{S(u^{(k)})}^2  \frac{b^{2(k-j)}}{\lambda_{n_k}^2}\\
& \ge \eta^2 
\sum_{k=0}^{j} \bignorm[\cf]{u^{(k)}}^2 
b^{2(k-j)}\\
& \ge  \eta^2 b^{2(k^{*}-j)} \qquad \text{ where } k^* = \argmax_{0 \le k \le j} \bignorm[\cf]{u^{(k)}} \\
& \ge \eta^2 = \frac{(a-1)^2c^2}{(a+1)^2R^2} \\
&
=\frac{(a-1)^2 \rho^2}{(a+1)^2A^2}\left[\left(1+\frac{(a-1)^2}{(a+1)^{2} A^2 }\right)\sum_{k=1}^j \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}\right]^{-1} \\
& = \rho^2 \left[\left\{\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right\}\sum_{k=1}^j \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}\right]^{-1} 
\end{align*}

This lower bound must be satisfied by $j$ to be consistent with the assumption $\comp(\ca(\cc),\varepsilon,\rho) \le n_j - 1$.  Thus, for any $j$ violating this inequality it follows that $\comp(\ca(\cc),\varepsilon,\rho) \ge n_j$.  This implication provides a lower bound on $\comp(\ca(\cc),\varepsilon,\rho)$.
\end{proof}

@@

\yuhannote{The figure is here in comment}
\begin{figure}
	\centering
	\includegraphics[width = 8 cm]{ProgramsImages/SimDirectSolErr.eps}
\end{figure}
